---
title: "Logistic Regression: Picking a Method"
subtitle: "stats::glm, rstanarm, or brms, for a dose-response fit"
description: >
  Comparing three ways to fit a logistic regression of dose against response:
  the frequentist stats::glm, and two Bayesian options, rstanarm and brms.
categories: [PMX guide, working document]
date: last-modified
execute:
  freeze: auto
  warning: false
  message: false
toc: true
toc-depth: 2
---

::: {.callout-note appearance="simple"}
**Draft, not fully worked through.** This compares three ways to fit the same
model rather than arguing for one. The `glm` fit below runs live; the two
Bayesian fits do not, because `rstanarm` and `brms` are not installed in the
environment that builds this site. Their code and Andy's original conclusions
are kept as a record of the comparison, not as a verified result.
:::

```{r setup}
library(dplyr)
library(ggplot2)
library(xgxr)

xgx_theme_set()
knitr::opts_chunk$set(fig.height = 4, fig.width = 5)
```

## 1. The Question

Fitting logistic regression for dose against both adverse event rates and
overall response rates is a common analysis in dose finding. This page asks
whether `stats::glm` is sufficient, or whether a Bayesian approach through
`rstanarm` or `brms` is preferable. The same data is fit with three different
functions:

1. `stats::glm` --- frequentist.
2. `rstanarm::stan_glm` --- Bayesian, a package built specifically for `glm`
   models.
3. `brms::brm` --- Bayesian, using a more flexible modeling package.

**Andy's conclusion from the original comparison:** all three gave similar
predictions, `stats::glm` ran fastest and needed no prior specification, so
for a quick, exploratory answer `stats::glm` is the simplest choice. Where
strong prior information exists, it should inform the analysis and a
Bayesian approach should be used. `rstanarm::stan_glm` runs faster than `brms::brm`,
because its models are precompiled, and is the more convenient of the two.

## 2. Caveat: The Limits of Logistic Regression

A binary endpoint carries less information than a continuous one, so a
logistic regression fit like the ones below is expected to be of limited
value on its own; an analysis with a continuous endpoint will usually be
needed to drive dosing decisions. [The sizing-studies
project](../sizing-studies/index.qmd) works this out in detail. Frank
Harrell's *Regression Modeling Strategies* gives the quick version: to
estimate a baseline rate to within ±10%, at least 96 patients are needed
([Section 10.2.2](https://hbiostat.org/rmsc/lrm.html)). At the best case,
$p = 0.5$, the variance of the estimate of $p$ is $p(1-p)/n$, giving a 95%
confidence interval of $1.96\sqrt{p(1-p)/n} = 0.1$ at $p = 0.5$ and $n = 96$.

## 3. The Data

96 patients from a synthetic Phase 1/2 oncology dataset: assigned dose,
baseline covariates, and RECIST-derived response. See [the project
index](index.qmd#data) for the full column list.

```{r data}
data <- readRDS("../../data/synthetic/dose-response/dose_tumor_resp.rds")

data |>
  group_by(DOSE) |>
  summarise(
    N = n(),
    N_responders = sum(OR == 1, na.rm = TRUE),
    Percent_responders = round(N_responders / N * 100)
  ) |>
  knitr::kable(caption = "Response by dose")
```

## 4. Frequentist: stats::glm

```{r logreg-glm}
start_time <- Sys.time()
model <- glm(OR ~ DOSE + ECOG0 + LD0, family = binomial, data = data)
run_time <- Sys.time() - start_time

model |>
  broom::tidy() |>
  arrange(p.value) |>
  select(-statistic) |>
  mutate(
    estimate  = signif(estimate, 2),
    std.error = signif(std.error, 2),
    p.value   = signif(p.value, 2)
  ) |>
  filter(term != "(Intercept)") |>
  knitr::kable(caption = "glm coefficient estimates")
```

`ECOG0` is baseline Eastern Cooperative Oncology Group (ECOG) performance
status (0 or 1) and `LD0` is the baseline sum of target-lesion diameters.

```{r logreg-glm-plot}
data_sim <- data.frame(
  DOSE = seq(0, 16, by = 0.1), ECOG0 = 1, LD0 = median(data$LD0)
)
p <- predict(model, data_sim, se.fit = TRUE)
f_trans <- function(x) exp(x) / (1 + exp(x))
data_sim$fit   <- f_trans(p$fit)
data_sim$lower <- f_trans(p$fit - 1.96 * p$se.fit)
data_sim$upper <- f_trans(p$fit + 1.96 * p$se.fit)

breaks <- c(0, 7, 11, 13, 16)
g_data <- ggplot(mapping = aes(x = DOSE, y = OR)) +
  geom_jitter(data = data, alpha = 0.5, width = 0, height = 0.1, size = 2) +
  xgx_stat_ci(data = data, distribution = "binomial", breaks = breaks, geom = "point", shape = 0, size = 5) +
  xgx_stat_ci(data = data, distribution = "binomial", breaks = breaks, geom = "errorbar", linewidth = 0.5) +
  scale_x_continuous(breaks = c(0, 4, 8, 12, 16), limits = c(0, 16)) +
  scale_y_continuous(breaks = c(0, 0.5, 1), labels = scales::percent) +
  labs(x = "Dose (mg)", y = "Probability of response")

g_data +
  geom_line(data = data_sim, aes(y = fit), linewidth = 1) +
  geom_ribbon(data = data_sim, aes(ymin = lower, ymax = upper, y = NULL), alpha = 0.25) +
  ggtitle(paste0("Frequentist stats::glm, run time = ", signif(run_time, 2), "s"))
```

## 5. Bayesian: rstanarm::stan_glm

`rstanarm` is not installed in the environment that builds this site, so this
chunk does not run. It is kept as the record of what was fit.

```{r rstanarm-code, eval = FALSE}
library(rstanarm)

fit_rstan <- stan_glm(
  OR ~ DOSE + LD0 + ECOG0,
  data = data,
  family = binomial(link = "logit"),
  prior = normal(c(-2, 0, 0), 3),
  prior_intercept = normal(0.5, 3),
  QR = TRUE,
  seed = 123,
  refresh = 0
)

data_sim <- tibble::tibble(DOSE = seq(0, 16, length.out = 100), ECOG0 = 0, LD0 = median(data$LD0))
p <- predict(fit_rstan, newdata = data_sim, se.fit = TRUE)
data_sim$fit   <- f_trans(p$fit)
data_sim$lower <- f_trans(p$fit - 1.96 * p$se.fit)
data_sim$upper <- f_trans(p$fit + 1.96 * p$se.fit)
```

## 6. Bayesian: brms::brm

`brms` is not installed either. Also kept as a record.

```{r brms-code, eval = FALSE}
library(brms)

model_blrm <- bf(
  OR ~ beta0 + exp(beta1) * DOSE / 16 + betatum0 * log(LD0) + betaecog * ECOG0,
  beta0 ~ 1, beta1 ~ 1, betatum0 ~ 1, betaecog ~ 1,
  family = bernoulli(), nl = TRUE
)

priors <- c(
  prior(prior = "normal(-2.20, 2)", class = "b", nlpar = "beta0"),
  prior(prior = "normal(-1.02, 1)", class = "b", nlpar = "beta1"),
  prior(prior = "normal(0, 5)",     class = "b", nlpar = "betatum0"),
  prior(prior = "normal(0, 5)",     class = "b", nlpar = "betaecog")
)

blrm_fit <- brm(
  model_blrm, data = data, iter = 7500, warmup = 2500,
  seed = 9999, chains = 4, prior = priors, cores = 4
)
```

## 7. Acknowledgements

Thanks to Juan Gonzalez-Maffe for the original `brms` code, and to Sebastian
Weber, Lukas Widner and Andrew Bean for help understanding these options.
