---
title: "Logistic Regression: Diagnostics"
subtitle: "Checking calibration, residuals and linearity, and the gaps left in the checklist"
description: >
  A minimal set of diagnostics for a logistic regression fit to dose-response
  data: calibration against observed rates, average residual versus outcome,
  and linearity of the continuous covariates.
categories: [PMX guide, working document]
date: last-modified
execute:
  freeze: auto
  warning: false
  message: false
toc: true
toc-depth: 2
---

::: {.callout-note appearance="simple"}
**Draft, with gaps.** [Section 5](#open) lists what this page does not yet
do: a Hosmer-Lemeshow test, an $R^2$ calculation, and a Brier score, among
others. What is here runs against real data.
:::

```{r setup}
library(dplyr)
library(ggplot2)
library(xgxr)

xgx_theme_set()
knitr::opts_chunk$set(fig.height = 4, fig.width = 5)
```

## 1. Purpose

A logistic regression of dose against adverse event rate or overall response
rate is a common analysis in dose finding. Given a fit, the next question is
whether it is adequate. This page gives a minimal set of diagnostics for
answering that.

Some of the original diagnostic work behind this page was developed by
Edward Waldron and a summer intern.

## 2. The Data and the Fit

The same 96-patient synthetic dataset as [Picking a
Method](picking-a-method.qmd); see [the project index](index.qmd#data) for
the column list.

```{r data}
data <- readRDS("../../data/synthetic/dose-response/dose_tumor_resp.rds") |>
  mutate(observed = OR)

data |>
  group_by(DOSE) |>
  summarise(
    N = n(),
    N_responders = sum(OR == 1, na.rm = TRUE),
    Percent_responders = round(N_responders / N * 100)
  ) |>
  knitr::kable(caption = "Response by dose")
```

```{r fit}
model <- glm(OR ~ DOSE + ECOG0 + LD0, family = binomial, data = data)
f_trans <- function(x) exp(x) / (1 + exp(x))
```

## 3. Check Calibration: Observed Versus Predicted Probability

This compares the observed and predicted probability of response within ten
bins of predicted probability. If the identity line (slope 1, intercept 0)
passes through most of the intervals' confidence bars, the fit's calibration
looks reasonable.

```{r calibration}
data_pred <- data |>
  mutate(
    pred = f_trans(predict(model)),
    group_id = Hmisc::cut2(pred, g = 10)
  )

data_plot <- data_pred |>
  group_by(group_id) |>
  summarise(
    total = n(),
    num_observed = sum(observed),
    mean_pred = mean(pred),
    mean_observed = mean(observed),
    upper_CI = binom::binom.confint(num_observed, total, conf.level = 0.95, method = "exact")$upper,
    lower_CI = binom::binom.confint(num_observed, total, conf.level = 0.95, method = "exact")$lower
  )

ggplot(data_plot, aes(x = mean_pred, y = mean_observed)) +
  geom_point() +
  geom_abline(slope = 1, intercept = 0, color = "blue") +
  geom_errorbar(aes(ymin = lower_CI, ymax = upper_CI), width = 0.005) +
  labs(x = "Predicted event rate", y = "Observed event rate")
```

A formal companion to this plot is the Hosmer-Lemeshow test, not yet
implemented here; see [Open questions](#open).

## 4. Check for Inadequacies: Average Residual Versus Outcome

The data is divided into bins by fitted value, with roughly equal patients
per bin, and the average residual is plotted against the average fitted
value for each bin. More bins show more local pattern in the residuals but
each needs enough points to keep the average from being noisy. About 95% of
binned residuals are expected to fall within $\pm 2$ standard errors, i.e.
$\pm 2\sqrt{p(1-p)/n}$ for $n$ points per bin.

```{r residuals}
n <- nrow(data)
nbins <- if (n >= 100) floor(sqrt(n)) else if (n > 10) 10 else floor(n / 2)

resid_bin <- data.frame(
  pred  = predict(model, type = "response"),
  resid = residuals(model)
) |>
  mutate(group_id = Hmisc::cut2(pred, g = nbins)) |>
  group_by(group_id) |>
  summarise(
    pred_mean  = mean(pred),
    resid_mean = mean(resid),
    resid_sd   = sd(resid),
    n          = length(pred),
    resid_upper =  1.96 * resid_sd / sqrt(n),
    resid_lower = -1.96 * resid_sd / sqrt(n)
  )

ggplot(resid_bin, aes(x = pred_mean, y = resid_mean)) +
  geom_point() +
  geom_ribbon(aes(ymin = resid_lower, ymax = resid_upper), alpha = 0.2) +
  geom_hline(yintercept = 0, color = "blue") +
  labs(x = "Average predicted response", y = "Average residual") +
  ggtitle(paste0("Average of ", signif(mean(resid_bin$n), 2), " data points per bin"))
```

## 5. Check the Linearity Assumption on Continuous Covariates

Logistic regression assumes a linear relationship between each continuous
covariate and the log-odds (logit) of the outcome. Plotting the logit
against each covariate, with a loess smooth, checks that visually.

```{r linearity}
data |>
  mutate(logit = predict(model)) |>
  tidyr::pivot_longer(cols = c(DOSE, LD0)) |>
  ggplot(aes(x = value, y = logit)) +
  geom_point() +
  geom_smooth(method = "loess") +
  facet_wrap(~name, scales = "free_x") +
  labs(x = "Predictor", y = "Logit")
```

## 6. Open Questions {#open}

Carried over from the original working notes, unresolved:

- **Hosmer-Lemeshow test.** Not implemented. `glmtoolbox::hltest` runs it,
  though that package may not be available in every R environment.
- **An $R^2$ measure.** Several candidates exist; see [Frank Harrell's
  discussion](https://hbiostat.org/bib/r2.html). The `rms` package's
  `R2Measures` function gives a choice of which to use.
- **Brier score**, covering both calibration and discrimination.
- **Other metrics** proposed in Harrell's lecture notes on binary logistic
  regression, Section 8-40.
- **The residual standard error** used in the Section 4 plot should be
  reviewed against a reference rather than taken as given.
