---
title: "A Model-Based Comparison for Three Cohorts of Ten"
subtitle: "Response rates, dose selection and evidence that the population meets the target"
description: "Reproducible binomial calculations and simulations comparing observed responder counts with continuous biomarker models, including misspecified dose-response and variance assumptions."
categories: [Dose-response methods, simulation]
date: last-modified
toc: true
toc-depth: 2
execute:
  echo: false
  warning: false
  message: false
---

```{r}
source("scripts/tdp-calculations.R")
validate_tdp_calculations()
benchmark <- run_tdp_benchmark()
pct <- function(x, digits = 1) paste0(formatC(100*x, format="f", digits=digits), "%")
value <- function(table, scenario, method, column) {
  table[table$scenario == scenario & table$method == method, column]
}
library(ggplot2)
theme_set(theme_minimal(base_size = 12))
```

**Continuous-data modeling can improve this decision, but the benefit depends
on the assumptions used to share information.** The calculations below keep
the response definition fixed: at least an 80% biomarker decline at a
prespecified assessment, with a target population response rate of 90%.

These are synthetic, single-assessment examples. They do not quantify the
additional benefit of pharmacokinetics (PK) or longitudinal pharmacodynamics
(PD). The [specification](specification.qmd) defines those comparisons;
the [literature review](literature-review.qmd) explains their published basis.

## 1. What Nine Responders Out of Ten Establishes

If patients independently respond with probability $p$, the probability that
the observed cohort passes is

$$
P(K\ge9\mid p)=10p^9(1-p)+p^{10},\qquad K\sim\mathrm{Binomial}(10,p).
$$

```{r}
ps <- c(.5, .7, .8, .85, .9, .95)
knitr::kable(data.frame(
  `True population response rate` = pct(ps),
  `Chance of at least 9/10 responders` = pct(binomial_pass(ps)),
  check.names = FALSE
))
```

A cohort with a true response rate of 80% passes the rule
`r pct(binomial_pass(.8))` of the time. At exactly the 90% population target,
the observed cohort fails `r pct(1-binomial_pass(.9))` of the time. These are
properties of the rule, before introducing any model.

When three independent cohorts all have a true response rate of 80%, at
least one passes with probability
$1-\{1-P(K\ge9\mid p=0.8)\}^3$ =
`r pct(1-(1-binomial_pass(.8))^3)`.
Selecting any passing dose therefore requires evaluating the whole selection
procedure, including the opportunity to pass at several doses.

```{r}
k <- c(9, 10)
knitr::kable(data.frame(
  `Observed responders` = paste0(k, "/10"),
  `One-sided 95% exact lower bound on p` = pct(exact_lower(k, 10)),
  `Pr(p >= 90% | data), uniform Beta(1,1) prior` =
    pct(pbeta(.9, k+1, 10-k+1, lower.tail=FALSE)),
  check.names = FALSE
))
```

The confidence bound and the Bayesian probability answer different questions;
the latter depends on the stated prior. Neither analysis of 9/10 establishes
the population target with high certainty. Even 10/10 cannot reject
$H_0:p\le0.9$ in a conventional, nonrandomized one-sided 5% exact binomial
test: its smallest attainable p-value is $0.9^{10}$ = `r pct(.9^10)`.
With all successes, at least `r ceiling(log(.05)/log(.9))` patients at one
dose are needed for the one-sided 95% exact lower bound to exceed 90%.
This is an all-success bound, not a powered sample size recommendation.

## 2. Estimating the Same Response Rate From Continuous Data

Define $Y=\log\{B(t^*)/B(0)\}$, where $B$ is a positive biomarker and
$t^*$ is the assessment time. An 80% decline corresponds to
$Y\le c=\log(0.2)$. Under the illustrative model
$Y\mid d\sim N\{\mu(d),\sigma(d)^2\}$,

$$
p(d)=\Phi\!\left(\frac{c-\mu(d)}{\sigma(d)}\right).
$$

The fitted mean and the spread both determine the responder fraction.
Estimating only the mean decline, or counting responders among fitted
individual means after shrinkage, does not calculate this population
quantity. The normal model here describes observed log ratios, so its
standard deviation (SD) includes all variation in that observed endpoint.

For a single dose, the large-sample variance of the plug-in estimate is
approximately

$$
\operatorname{Var}(\widehat p_{\rm continuous})
\approx\frac{\phi(z)^2}{n}\left(1+\frac{z^2}{2}\right),
\quad z=\Phi^{-1}(p).
$$

The $z^2/2$ term accounts for estimating the standard deviation. It follows
by applying the delta method to the sample mean and standard deviation of
normal data. The binary estimator has variance $p(1-p)/n$.
The ratio of these variances is an approximate ratio of required sample
sizes for equally precise estimation of $p$.

```{r}
eff <- normal_efficiency(c(.5, .8, .9, .95))
knitr::kable(data.frame(
  `True response rate` = pct(eff$p),
  `Binary / continuous variance, SD estimated` = round(eff$estimated_sd, 2),
  `Binary / continuous variance, SD known` = round(eff$known_sd, 2),
  check.names = FALSE
))
```

At $p=0.9$, the estimated-SD calculation gives an efficiency factor of
`r round(normal_efficiency(.9)$estimated_sd, 2)`, equivalent asymptotically
to a `r pct(1-1/normal_efficiency(.9)$estimated_sd)` sample size reduction
for this estimation problem. Treating the SD as known would exaggerate the
gain when it must be learned from these patients. These calculations are
neither a finite-sample power result nor an additional benefit from PK.
The distributional-estimation idea goes back at least to
[Suissa (1991)](https://pubmed.ncbi.nlm.nih.gov/1999683/); the calculation
above is independently implemented for this example.

## 3. Comparing Dose Decisions

Each simulated study has ten independent patients at each of three relative
doses, 1, 2 and 4. There is one observed log biomarker ratio per patient,
complete follow-up, and no assay censoring. All methods receive the same
patients. The decision is to select the lowest tested dose whose estimated
response probability reaches 90%, or select no dose if none qualifies.

| Method | Information and assumptions |
|---|---|
| M0 Counts | Separate responder proportions. A dose passes at 9/10 or 10/10. |
| M1 Binary monotone | Isotonic binomial estimates: pool neighboring arms only when their observed response rates decrease with dose. |
| M2 Continuous per arm | Separate normal mean and SD for each dose's log ratio. |
| M3 Shared SD | Separate means, with a common residual SD estimated from all three arms. |
| M4 Dose curve | A straight line for mean log ratio against log dose, with a common residual SD. |

M2 isolates the gain from retaining response magnitudes. M3 shares variance
information. M4 also shares information about the means. With equal cohort
sizes, isotonic least squares on the observed proportions gives the
order-restricted binomial maximum likelihood estimates used in M1.
M4's slope is unconstrained; a reversal can be fitted. Its log-dose shape
only applies over the tested range and is not a mechanistic dose model.

```{r}
truth <- benchmark$truth
truth$scenario <- factor(truth$scenario, levels = unique(truth$scenario))
truth$label <- paste0(pct(truth$p), " (SD ", truth$sigma, ")")
truth_wide <- reshape(truth[,c("scenario","dose","label")],
                      idvar="scenario", timevar="dose", direction="wide")
names(truth_wide) <- c("Scenario", "Dose 1: p (SD)", "Dose 2: p (SD)", "Dose 4: p (SD)")
knitr::kable(truth_wide, row.names=FALSE)
```

The first two scenarios satisfy M4's shape and common-SD assumptions. The
plateau also satisfies the linear model, with zero slope. The late-rise
scenario violates the linear mean shape. The last scenario violates both
M4's mean shape and the shared-SD assumption; all doses are inadequate.
M2 remains correctly specified in every scenario because the log ratios
within each arm are normal.

```{r}
#| label: fig-selection
#| fig-width: 9
#| fig-height: 8
#| fig-cap: "Point-estimate decisions across 10,000 simulated studies per scenario. Correct means the lowest adequate dose, or no dose when all are inadequate. Selecting an inadequate dose means selecting one with true p below 90%. Horizontal bars are Monte Carlo 95% intervals, not uncertainty in the assumed biological scenarios."
tab <- benchmark$point
chart <- rbind(
  data.frame(tab[,c("scenario","method")], outcome="Correct decision",
             probability=tab$correct, mcse=tab$correct_mcse),
  data.frame(tab[,c("scenario","method")], outcome="Inadequate dose selected",
             probability=tab$inadequate, mcse=tab$inadequate_mcse)
)
chart$scenario <- factor(chart$scenario, levels=unique(tab$scenario))
chart$method <- factor(chart$method, levels=rev(unique(tab$method)))
ggplot(chart, aes(probability, method, color=outcome)) +
  geom_point(position=position_dodge(width=.45), size=2) +
  geom_errorbar(aes(xmin=pmax(0,probability-1.96*mcse),
                    xmax=pmin(1,probability+1.96*mcse)),
                orientation="y", position=position_dodge(width=.45), width=.15) +
  facet_wrap(~scenario, ncol=2) +
  scale_x_continuous(limits=c(0,1), labels=function(x) paste0(100*x,"%")) +
  scale_color_manual(values=c("Correct decision"="#176B63", "Inadequate dose selected"="#B94C35")) +
  labs(x="Fraction of simulated studies", y=NULL, color=NULL) +
  theme(legend.position="bottom", panel.grid.major.y=element_blank())
```

When only the high dose qualifies under a smooth curve, correct selection
rises from
`r pct(value(tab,"Smooth: high qualifies","M0 Counts","correct"))` with
counts to
`r pct(value(tab,"Smooth: high qualifies","M4 Dose curve","correct"))`
with the dose curve. In the late-rise scenario, the shared-SD model with
separate means does better than imposing the linear mean shape.
In the unequal-SD scenario, variance pooling makes the decision worse.

The point-estimate rules do not control a 5% false-positive rate. Their
different false-positive rates also mean that the difference in correct
selection cannot be called a gain in statistical power at equal type I
error. Precision, decision accuracy and evidence against a null hypothesis
are separate comparisons.

::: {.callout-note collapse="true" title="All Point-Estimate Decisions"}

```{r}
display_decisions <- function(a) {
  data.frame(Scenario=a$scenario, Method=a$method,
    `Correct decision`=pct(a$correct),
    `Inadequate dose`=pct(a$inadequate),
    `Higher adequate dose`=pct(a$higher_adequate),
    `No dose`=pct(a$no_dose), check.names=FALSE)
}
knitr::kable(display_decisions(tab), row.names=FALSE)
```

When all doses are inadequate, a correct decision and a no-dose decision
are the same event. Those two columns therefore overlap in those scenarios.
In scenarios with an adequate dose, the four columns partition the decisions.

:::

## 4. Requiring Evidence That a Dose Exceeds the Target

A second comparison asks for evidence against $H_0:p(d)\le0.9$.
Every method receives a one-sided error allowance of $0.05/3$ per tested
dose. The Bonferroni bound limits the chance of declaring any truly
inadequate dose to at most 5%, provided each model's test is valid.
The lowest dose that rejects its null is selected.

For normal data with fitted mean $\widehat\mu_d$, residual SD $s$,
residual degrees of freedom $\nu$, and mean-variance factor
$h_d=\operatorname{Var}(\widehat\mu_d)/\sigma^2$,

$$
T_d=\frac{c-\widehat\mu_d}{s\sqrt{h_d}}
\sim t_\nu\!\left(\frac{z_d}{\sqrt{h_d}}\right),
\qquad z_d=\Phi^{-1}\{p(d)\}.
$$

The noncentral-$t$ distribution gives an exact upper critical value at
$z_d=\Phi^{-1}(0.9)$. This accounts for uncertainty in the fitted mean and
SD. M2 uses $h_d=1/10$, $\nu=9$; M3 uses $h_d=1/10$, $\nu=27$; M4 uses
the regression prediction variance and $\nu=28$.
The binary comparator uses an exact binomial test. M1 is omitted from this
comparison because an order-restricted testing procedure has not been
implemented.

```{r}
cert <- benchmark$certified
knitr::kable(display_decisions(cert), row.names=FALSE)
```

The binary test cannot pass with ten patients per dose, as Section 1 showed.
The normal-data tests can pass because they also observe how far the
responses lie beyond the threshold. Their power is still limited close to
the population target. A more distant, clearly adequate high dose is easier
to certify than the lowest adequate dose.

The apparent certainty can fail under misspecification. In the unequal-SD
scenario, M3 selects an inadequate dose in
`r pct(value(cert,"Unequal SD: none qualifies","M3 Shared SD","inadequate"))`
of studies despite its nominal 5% familywise allowance. M2's corresponding
rate is
`r pct(value(cert,"Unequal SD: none qualifies","M2 Continuous per arm","inadequate"))`.
This is why an integrated model needs a calibrated decision rule and
plausible misspecification scenarios before a claim about sample savings.

## 5. Reproduction and Next Comparison

The [R source](scripts/tdp-calculations.R) contains the settings, generators,
estimators, decision rules and numerical checks. This page reruns the
calculations at render time. The simulation uses
`r format(tdp_settings$nsim, big.mark=",")` replicates per scenario and
seed `r tdp_settings$seed`. At that simulation size, the largest possible
Monte Carlo standard error for a reported proportion is
`r formatC(100*sqrt(.25/tdp_settings$nsim), format="f", digits=2)` percentage
points; uncertainty about the scenario
parameters is not included.

The run also returns dose-specific bias and root mean squared error in
`benchmark$estimates`. The analytical checks verify the binomial tail,
confidence-bound calculation, no-dose selection and the noncentral-$t$
test's boundary rejection rate by numerical integration.

The next comparison adds repeated biomarker measurements, followed by PK,
using the [same target and decision rules](specification.qmd#integrated-model).
Those additions must improve on M2–M4 to establish an incremental benefit
from integrated pharmacometric modeling.
