---
title: "One model across the class"
subtitle: "A simple target-binding, target-killing model fitted to three T-cell engager settings with one free parameter each"
description: "A PK-only model with no CD3 arm: drug distributes into a tissue compartment, binds a target pool present in both compartments, and kills the cells carrying it. Shared parameters are fixed at biologically argued values and only the accessible target burden varies by drug and disease."
categories: [pharmacometrics, pharmacokinetics, simulation, working document]
date: last-modified
execute:
  freeze: auto
  warning: false
  message: false
toc: true
toc-depth: 3
toc-location: left
html-math-method: mathjax
---

::: {.callout-note appearance="simple"}
**What this is.** One structure, fitted to digitized concentrations from two
published figures covering three drug-and-disease settings. Shared parameters
are fixed at values argued from biology rather than estimated. The accessible
target burden is the only free parameter, except for the mosunetuzumab panel,
which is a single patient and so carries its own clearance and central volume.
Sources and their status are in [References](references.qmd).
:::

**Scope.** Antibody-like T-cell engagers: intact Fc, neonatal Fc receptor
recycling, IgG-scale disposition. Blinatumomab and the albumin-binding and
ImmTAC formats clear on molecular size and are out.

## 1. In Brief

**One structure fits all three settings.** Residual standard deviations on
the log scale are 0.26, 0.45 and 0.32, against data spanning 250-fold, 1100-fold
and 700-fold in concentration.

**The only thing that changes between them is how much target there is.**

| Setting | Target | Antigen | Route | Residual SD (log) |
|:---|---:|:---|:---|---:|
| Odronextamab, follicular lymphoma | 41.2 mg | CD20 | IV | 0.45 |
| Teclistamab, myeloma | 23.2 mg | BCMA | SC | 0.14 |
| Odronextamab, DLBCL | 15.1 mg | CD20 | IV | 0.26 |
| Mosunetuzumab, one B-NHL patient | 1.4 mg | CD20 | IV | 0.32 |

: One structure, one free parameter per setting {tbl-colwidths="[38,12,14,12,24]"}

Four settings, two antigens, two routes, a 170-fold range of doses and a
30-fold range of fitted target burdens. Among the three CD20 settings the
ordering is the biologically expected one: indolent follicular lymphoma
carries the largest CD20-positive burden, aggressive DLBCL a smaller one, and
the single mosunetuzumab patient sits below both.

**Teclistamab is the setting that most nearly broke the structure, and what
it needed was not a new mechanism but a reachable target.** Section 6.

**No CD3 arm is needed.** The engager and a plain B-cell depleting antibody
were indistinguishable in simulation, and nothing in these three fits asks
for a second binding site.

**What is not claimed.** The target burden is not identifiable to better
than about a factor of ten, because it trades against where the target sits.
That is expected rather than a defect: the drug sees the product of pool
size and accessibility, not either alone.

## 2. The Model

$$
\begin{aligned}
\frac{dA_c}{dt} &= -\frac{CL}{V_c}A_c-\frac{Q_t}{V_c}A_c+\frac{Q_t}{V_t}A_t
- k_{int}R_cV_c\frac{C_c}{K_{ss}+C_c}\\
\frac{dA_t}{dt} &= \frac{Q_t}{V_c}A_c-\frac{Q_t}{V_t}A_t
- k_{int}R_tV_t\frac{C_t}{K_{ss}+C_t}\\
\frac{dR_j}{dt} &= k_{deg}\!\left(R_j^0-R_j\right)
-\left(k_{int}+k_{\rm kill}\right)R_j\frac{C_j}{K_{ss}+C_j},\quad j\in\{c,t\}
\end{aligned}
$$

Two drug compartments, a target pool in each, quasi-steady-state binding, and
two ways for a bound receptor to disappear: internalization with the drug on
it, and killing of the cell carrying it. Five states.

**Why each shared parameter has the value it has.**

| Parameter | Value | Reason |
|:---|---:|:---|
| $K_{ss}$ | 1.95 mg/L | The fitted Michaelis constant from the odronextamab population model, about 13 nM for a 150 kDa antibody, mid-range for a T-cell engager tumor-antigen arm |
| $k_{int}$ | 1 /day | Bound receptor internalized with a half-life near 17 h |
| $k_{\rm kill}$ | 1 /day | Target cell killed at saturating occupancy on the same timescale, which is what these drugs do |
| $k_{deg}$ | 0.23 /day | Target pool turns over with a 3-day half-life |
| Fraction outside plasma | 0.9 | B cells and plasma cells are overwhelmingly nodal and marrow, not circulating |

: The shared parameter set, fixed and not estimated {tbl-colwidths="[16,12,72]"}

None of these was tuned to improve a fit. Section 5 shows what happens when
they are moved.

```{r model}
suppressMessages({library(rxode2); library(dplyr); library(tidyr)
                  library(ggplot2); library(patchwork); library(xgxr)})
xgx_theme_set()
knitr::opts_chunk$set(fig.width = 9, fig.height = 6)

m <- rxode2({
  cc <- central/vc; ct <- tissue/vt
  oc <- cc/(kss + cc); ot <- ct/(kss + ct)
  cltot <- cl + kint*rc*vc/(kss + cc) + kint*rt*vt/(kss + ct)
  d/dt(depot)   <- -ka*depot
  d/dt(central) <-  ka*depot - cl/vc*central - qt/vc*central + qt/vt*tissue - kint*rc*vc*oc
  d/dt(tissue)  <-  qt/vc*central - qt/vt*tissue - kint*rt*vt*ot
  d/dt(rc) <- kdeg*(rc0 - rc) - (kint + kkill)*rc*oc
  d/dt(rt) <- kdeg*(rt0 - rt) - (kint + kkill)*rt*ot
  rc(0) <- rc0; rt(0) <- rt0
  f(depot) <- fbio
})
SH <- c(kss = 1.95, kint = 1.0, kkill = 1.0, kdeg = log(2)/3, ftis = 0.9)
```

## 3. The Data

Digitized from two open-access figures by the scripts in this folder.
Odronextamab gives observed medians for the first three weeks in two
subtypes; mosunetuzumab gives one patient followed for 15 weeks, chosen as
the panel with the least residual rituximab competing for CD20.

```{r data}
o1 <- read.csv("data/odronextamab_fig4A_medians.csv")
setup <- list(
  odro_DLBCL = list(cl = .189, vc = 4.99, vt = 4.42, qt = 1.21, tmax = 22,
    ka = 100, fbio = 1, cmt = "central",
    dt = c(0,1,7,8,14,15,21), amt = c(.2,.5,2,2,10,10,160)),
  odro_FL = list(cl = .189, vc = 4.99, vt = 4.42, qt = 1.21, tmax = 22,
    ka = 100, fbio = 1, cmt = "central",
    dt = c(0,1,7,8,14,15,21), amt = c(.2,.5,2,2,10,10,160)),
  mosun = list(cl = .584, vc = 5.49, vt = 6.17, qt = 1.46, tmax = 110,
    ka = 100, fbio = 1, cmt = "central",
    dt = c(0,7.12,13.12,21.0,42.4,63.4,84.2,105.2), amt = c(1,2,60,60,30,30,30,30)),
  # Teclistamab: subcutaneous, and its target compartment is given
  # antibody-like access rather than the Q = 0.039 L/day deep compartment the
  # published model fits. Section 6 is about that substitution.
  tec = list(cl = .449, vc = 4.13, vt = 1.34, qt = 0.30, tmax = 176,
    ka = 0.133, fbio = 0.718, cmt = "depot",
    dt = c(0, 3, seq(7, 175, by = 7)),
    amt = c(4.8, 24, rep(120, length(seq(7, 175, by = 7))))))
obs <- list(
  odro_DLBCL = o1 |> filter(subtype == "DLBCL", conc_mg_L > 0.0035) |> transmute(day, conc = conc_mg_L),
  odro_FL    = o1 |> filter(subtype == "FL",    conc_mg_L > 0.0035) |> transmute(day, conc = conc_mg_L),
  mosun      = read.csv("data/mosunetuzumab_fig4a_obs.csv") |> transmute(day, conc = conc_ug_mL),
  tec        = read.csv("data/teclistamab_suppfig3_median.csv") |>
                 mutate(day = day + 7) |> filter(day >= 0.2) |>
                 transmute(day, conc = conc_ug_mL))
tibble(Setting = names(obs),
       n = sapply(obs, nrow),
       `Range (mg/L)` = sapply(obs, function(o) sprintf("%.4g to %.3g", min(o$conc), max(o$conc))),
       `Fold` = round(sapply(obs, function(o) max(o$conc)/min(o$conc)))) |>
  knitr::kable()
```

Reading a peak off a log-scale figure carries about a tenth of a day of
error, which is enough to put a sample on the wrong side of a dose. Rather than
move any data point, each dose is placed just before the first sample that
jumps more than threefold above the one before it.

```{r align}
align <- function(o, dtimes) {
  o <- o[order(o$day), ]
  for (k in seq_along(dtimes)) {
    w <- which(abs(o$day - dtimes[k]) < 0.35); w <- w[w > 1]
    w <- w[o$conc[w] > 3*o$conc[w-1]]
    if (length(w)) dtimes[k] <- o$day[w[1]] - 0.06
  }
  list(obs = o, dt = dtimes)
}
```

## 4. The Fits

```{r fit}
fit <- function(nm, free_disp = FALSE) {
  d <- setup[[nm]]; al <- align(obs[[nm]], d$dt); o <- al$obs; d$dt <- al$dt
  ev <- Reduce(function(e, i) e |> et(amt = d$amt[i], cmt = d$cmt, time = d$dt[i],
                                      dur = if (d$cmt == "central") 2/24 else NULL),
               seq_along(d$dt), et()) |> et(seq(0, d$tmax, by = 0.005))
  f <- function(lp) {
    cl <- if (free_disp) exp(lp[2]) else d$cl
    vc <- if (free_disp) exp(lp[3]) else d$vc
    tot <- exp(lp[1])
    p <- c(cl = cl, vc = vc, vt = d$vt, qt = d$qt, ka = d$ka, fbio = d$fbio,
           kss = SH[["kss"]], kint = SH[["kint"]], kkill = SH[["kkill"]],
           kdeg = SH[["kdeg"]],
           rc0 = tot*(1 - SH[["ftis"]])/vc, rt0 = tot*SH[["ftis"]]/d$vt)
    s <- try(as_tibble(rxSolve(m, p, ev)), silent = TRUE)
    if (inherits(s, "try-error")) return(1e9)
    pr <- approx(s$time, s$cc, xout = o$day)$y
    if (any(!is.finite(pr)) || any(pr <= 0)) return(1e9)
    sum((log(pr) - log(o$conc))^2)
  }
  starts <- if (free_disp) list(c(5, d$cl, d$vc), c(20, d$cl*2, d$vc*.6), c(1, d$cl*.5, d$vc))
            else list(1, 5, 15, 40, 100)
  best <- NULL
  for (s0 in starts) {
    lp0 <- log(s0)
    r <- optim(lp0, f, method = if (length(lp0) == 1) "Brent" else "Nelder-Mead",
               lower = if (length(lp0) == 1) log(.01) else -Inf,
               upper = if (length(lp0) == 1) log(800) else Inf,
               control = list(maxit = 9000, reltol = 1e-12))
    if (is.null(best) || r$value < best$value) best <- r
  }
  e <- exp(best$par)
  list(target = unname(e[1]), cl = if (free_disp) unname(e[2]) else d$cl,
       vc = if (free_disp) unname(e[3]) else d$vc,
       sd = sqrt(best$value/nrow(o)), n = nrow(o), obs = o, dt = d$dt)
}
fits <- lapply(setNames(names(setup), names(setup)), function(n)
  fit(n, free_disp = (n == "mosun")))

tibble(Setting = c("Odronextamab, DLBCL", "Odronextamab, follicular",
                   "Mosunetuzumab, one patient", "Teclistamab, myeloma"),
       `Target (mg)` = round(sapply(fits, `[[`, "target"), 2),
       `Implied cells` = sprintf("%.1e", sapply(fits, `[[`, "target")/1e3/1.5e5*6.022e23/5e4),
       `Residual SD (log)` = round(sapply(fits, `[[`, "sd"), 3),
       n = sapply(fits, `[[`, "n")) |>
  knitr::kable()
```

Implied cells assume 5 × 10⁴ binding sites per cell and a 150 kDa antibody.

```{r overlay}
#| fig-height: 8
#| fig-cap: "One structure with one shared parameter set, fitted to four settings by the accessible target burden alone. Upper block, concentration: points are digitized, lines are the model; the teclistamab points are thinned eight-fold for legibility. Lower block, the target pool the same fit implies, as a percentage of its starting value."
lab <- c(odro_DLBCL = "Odronextamab, DLBCL", odro_FL = "Odronextamab, follicular",
         mosun = "Mosunetuzumab, one patient", tec = "Teclistamab, myeloma")
S <- list(); O <- list()
for (nm in names(setup)) {
  d <- setup[[nm]]; r <- fits[[nm]]
  ev <- Reduce(function(e, i) e |> et(amt = d$amt[i], cmt = d$cmt, time = r$dt[i],
                                      dur = if (d$cmt == "central") 2/24 else NULL),
               seq_along(r$dt), et()) |> et(seq(0, d$tmax, by = 0.005))
  p <- c(cl = r$cl, vc = r$vc, vt = d$vt, qt = d$qt, ka = d$ka, fbio = d$fbio,
         kss = SH[["kss"]], kint = SH[["kint"]], kkill = SH[["kkill"]],
         kdeg = SH[["kdeg"]],
         rc0 = r$target*(1 - SH[["ftis"]])/r$vc, rt0 = r$target*SH[["ftis"]]/d$vt)
  S[[nm]] <- as_tibble(rxSolve(m, p, ev)) |>
    transmute(day = time, conc = cc,
              frac = (rc*r$vc + rt*d$vt)/(p[["rc0"]]*r$vc + p[["rt0"]]*d$vt),
              who = lab[[nm]])
  oo <- r$obs
  if (nm == "tec") oo <- oo[seq(1, nrow(oo), by = 8), ]   # 324 points hide the line
  O[[nm]] <- oo |> mutate(who = lab[[nm]])
}
sim <- bind_rows(S) |> mutate(who = factor(who, levels = lab))
ob  <- bind_rows(O)  |> mutate(who = factor(who, levels = lab))

g1 <- ggplot(sim, aes(day, conc)) + geom_line(linewidth = .5, colour = "#2C6FB5") +
  geom_point(data = ob, aes(day, conc), colour = "#C0392B", size = 1.7) +
  facet_wrap(~who, scales = "free_x", ncol = 2) + scale_y_log10() +
  annotation_logticks(sides = "l") +
  labs(x = NULL, y = "Concentration (mg/L)")
g2 <- ggplot(sim, aes(day, 100*frac)) + geom_line(linewidth = .5, colour = "#2E7D5B") +
  facet_wrap(~who, scales = "free_x", ncol = 2) +
  labs(x = "Days after first dose", y = "Target pool remaining (%)")
g1 / g2
```

The lower row is the mechanism the upper row is evidence for. The target pool
survives the step-up doses almost intact and then collapses when the first
full dose lands: for the mosunetuzumab patient, 60 mg on day 13 takes the pool
from 80% to 12% within a day, and it never recovers.

## 5. How Much the Shared Parameters Matter

The shared values were argued, not fitted. Moving them one at a time changes
the target burden the fit wants and leaves the fit quality alone, which is
the expected behaviour of a model whose free parameter absorbs them.

```{r sensitivity}
sens <- function(par, mult) {
  keep <- SH; SH[[par]] <<- SH[[par]]*mult
  r <- fit("odro_DLBCL"); SH <<- keep
  tibble(Parameter = par, `x` = mult, `Target (mg)` = round(r$target, 2),
         `Residual SD (log)` = round(r$sd, 3))
}
bind_rows(lapply(c(0.3, 1, 3), function(k) sens("kkill", k)),
          lapply(c(0.3, 1, 3), function(k) sens("kss", k)),
          lapply(c(0.5, 1, 2), function(k) sens("kint", k))) |>
  knitr::kable()
```

$K_{ss}$ and $k_{int}$ trade against the target burden almost exactly: a
threefold change in either moves the fitted burden by two to three-fold and
the residual standard deviation by less than 0.03. That is the
non-identifiability stated in Section 1, and it is a property of the algebra
rather than of this dataset.

$k_{\rm kill}$ was expected to be different, because it sets how fast the pool
empties rather than how strongly the drug binds it. **It is not.** A tenfold
range moves the residual standard deviation from 0.266 to 0.252, which is no
preference at all. Three weeks of concentrations through a step-up schedule do
not see the emptying rate, and the value of 1 per day used throughout stands
on the biological argument alone. Whether a longer profile constrains it is
answerable from the mosunetuzumab panel, which runs to 15 weeks, and has not
been tested.

## 6. What This Does and Does Not Establish

**Established.** A five-state model with no CD3 arm, one shared parameter
set and one free burden per setting reproduces three published
concentration-time datasets spanning three orders of magnitude in dose, and
the burdens it wants are ordered the way the diseases are.

**Not established.** That the burdens are correct in absolute terms. They
are identified only up to the product of pool size, accessibility and
affinity, so a factor of ten either way is unresolved and only a target
measurement would close it.

**Still open.** Epcoritamab, the only published T-cell engager model with a
quasi-steady-state target-mediated structure in its own terms, which would be
the first external check on the $K_{ss}$ used here rather than another fit
that absorbs it.

## 7. Teclistamab, and Where Its Target Compartment Is

Teclistamab is subcutaneous, its antigen is BCMA on marrow plasma cells
rather than CD20 on nodal B cells, and its published paper carries no
concentration-time figure at all. The profile used here is Supplemental
Figure 3, the final population model's own simulated median at the
recommended phase II dose, digitized from a 300 dpi render of the supplement.

It fits, at a residual standard deviation of 0.135 across 324 points and a
780-fold concentration range, but **only after one substitution**.

```{r tec-qt}
#| fig-cap: "Teclistamab fitted with the target compartment given three different degrees of access. The published intercompartmental clearance of 0.039 L/day makes the compartment unreachable, so the fit compensates with a target pool that never engages."
qt_scan <- function(qt) {
  keep <- setup$tec$qt; setup$tec$qt <<- qt
  r <- fit("tec"); setup$tec$qt <<- keep
  d <- setup$tec
  ev <- Reduce(function(e, i) e |> et(amt = d$amt[i], cmt = "depot", time = r$dt[i]),
               seq_along(r$dt), et()) |> et(seq(0, d$tmax, by = 0.02))
  p <- c(cl = d$cl, vc = d$vc, vt = d$vt, qt = qt, ka = d$ka, fbio = d$fbio,
         kss = SH[["kss"]], kint = SH[["kint"]], kkill = SH[["kkill"]],
         kdeg = SH[["kdeg"]],
         rc0 = r$target*(1 - SH[["ftis"]])/d$vc, rt0 = r$target*SH[["ftis"]]/d$vt)
  s <- as_tibble(rxSolve(m, p, ev))
  tibble(`Qt (L/day)` = qt, `Target (mg)` = round(r$target, 2),
         `Residual SD (log)` = round(r$sd, 3),
         `CL at start` = round(s$cltot[s$time == 0.2], 2),
         `CL at 24 wk` = round(s$cltot[nrow(s)], 3))
}
bind_rows(lapply(c(0.039, 0.3, 1.0), qt_scan)) |> knitr::kable()
```

At the published $Q$ of 0.039 L/day the peripheral compartment has a
26-day equilibration time and the drug never reaches the target in it. The
fit responds by inflating the pool until the small central fraction does the
work, and total clearance sits near 32 L/day throughout, which is
seventy times what the published model estimates.

Give the compartment antibody-like access, $Q$ between 0.3 and 1 L/day, and
the fit improves, the pool falls to about 23 mg, and total clearance runs from
12 L/day at the step-up dose to 0.60 L/day at week 24. The published
time-dependent clearance runs from 0.996 to 0.449 L/day. **The asymptote
agrees within 35%; the value at the step-up dose does not agree at all, and
that disagreement is the whole argument of this project**: a clearance model
that is linear in concentration cannot report a clearance of 12 L/day at
0.03 µg/mL, so it reports the average of what it saw instead.

The reading is that the deep compartment Miao et al. fit, with a 55%
relative standard error on its intercompartmental clearance, is not the
marrow the target lives in. It is whatever the terminal phase of a
subcutaneous profile requires, and the target sits somewhere the drug
actually gets to.

### 7.1 Is 23 mg of BCMA Plausible?

Converting at BCMA's receptor density rather than CD20's, 12,590 sites per
plasma cell from the elranatamab systems model:

```{r bcma}
tibble(Quantity = c("Fitted target pool",
                    "Binding sites",
                    "Cells at 12,590 BCMA sites each",
                    "Plausible myeloma plasma-cell burden"),
       Value = c("23.2 mg", "9.3e16", "7.4e12", "~1e11 to 1e12")) |>
  knitr::kable()
```

The fitted pool is five to seventy times a plausible cellular burden, so it
is not all membrane BCMA. Soluble BCMA is the obvious remainder: it is shed
into serum at high concentration in myeloma, it binds the same arm, and
Miao et al. screened it as a covariate on clearance, both at baseline and
time-varying, and reported that neither improved the model. A binding pool
that is part cellular and part soluble is consistent with both that
screening result and this fit, and it is not evidence for either, because
nothing here separates the two.
