---
title: "Why target-mediated drug disposition is rarely visible in T-cell engager pharmacokinetics"
subtitle: "Working specification"
description: "T-cell engagers are dosed far below the amount needed to saturate CD3, so a sharp concentration drop should be the usual sight and is not. This works through four explanations, tests each against published population pharmacokinetic models, and says which one carries the weight."
categories: [pharmacometrics, pharmacokinetics, oncology, working document]
date: last-modified
execute:
  freeze: auto
  warning: false
  message: false
toc: true
toc-depth: 3
toc-location: left
toc-expand: true
html-math-method: mathjax
---

::: {.callout-important appearance="simple"}
**Two sources have been read in full, and none of the transcribed parameter
values has been checked against a printed table.** The odronextamab,
mosunetuzumab, teclistamab and elranatamab parameters driving every figure and
calculation below were taken from an open-access full text and from
`nlmixr2lib` model files, not from the published tables. Sections 3 and 7 compute their conclusions
from those numbers, so both conclusions inherit whatever is wrong in them.
Section 5 rests on two drugs, which is where its own limits section says
it rests.
[References](references.qmd) carries a status marker on every source and
[the project index](index.qmd) lists the other documents in this folder.
:::

## In brief

**The observation.** A T-cell engager (TCE) binds CD3 on T cells with
nanomolar affinity, and CD3 internalizes with a half-life near an hour.
A first dose of 0.16 mg of epcoritamab or 0.7 mg of odronextamab is the
same order of magnitude as the CD3 receptors circulating in blood, and a
small fraction of what those receptors internalize in a week.
Every ingredient of textbook target-mediated drug disposition (TMDD) is
present, so the concentration-time profile after a priming dose should
fall off a cliff. Published profiles do not. The approved hematology
TCEs report terminal half-lives of 6 to 22 days, which is antibody-like.

**The answer this document argues for, in one line.** TMDD is there and is
large, it is reported as *time-dependent clearance* rather than as a
concentration nonlinearity, and the priming doses that could show the
nonlinearity sit below the Michaelis constant, where a Michaelis-Menten
elimination is arithmetically indistinguishable from a first-order one.

**The decisive number.** For odronextamab the fitted Michaelis constant is
$k_m$ = 1.95 mg/L and the 0.7 mg priming dose reaches a peak concentration
of 0.14 mg/L. The entire priming-dose profile lives at $C/k_m \le 0.07$.
Section 3 simulates that profile with target depletion switched off and
finds total clearance varying from 7.2 to 7.7 L/day across a 900-fold
concentration range. The mechanism is saturable and behaves linearly,
because the dose is too small to saturate it.

**What happens to the target sink.** It is destroyed. Four of the five
published population pharmacokinetic models for approved TCEs carry a
clearance term that decays with time since first dose, and the decay is
attributed to depletion of the target-bearing cells. Odronextamab's total
clearance falls from about 5.6 L/day to 0.22 L/day, a factor of 25.
Section 4 tabulates all six published models.

**Andy's flip-flop hypothesis, tested.** True for teclistamab early in
treatment, false for elranatamab, and inapplicable to the intravenous
drugs. Teclistamab's absorption half-life is 5.2 days against an effective
elimination half-life of 3.8 days at treatment start, so absorption is
rate-limiting then, and 8.4 days at steady state, so it is not
rate-limiting later. Elranatamab absorbs with a 4.7-day half-life and
eliminates with a 25-day one, so absorption is never rate-limiting for it.
Section 5 works both.

**Out of scope, stated once.**

- ❌ **A new TMDD model.** Nothing here is fit. Every parameter is
  transcribed from a published model and simulated forward.
- ❌ **Quantitative systems pharmacology.** No trimer formation, no
  cytokine cascade, no tumor dynamics. The CD3-plus-antigen binding model
  of Pearce et al. is the deepest mechanism used.
- ❌ **Dose selection.** No regimen, step-up schedule or starting dose
  should be taken from this document.
- ✅ **The parameter survey.** Compiling fixed and random effects across
  the published TCE models is a deliverable in its own right, and it lives
  in [PK parameters](pk-parameters.qmd) rather than here.

## 1. The Observation, as Numbers

Pearce et al. tabulate published half-lives for 29 TCEs in cynomolgus
monkey and in patients. The approved hematology molecules cluster near a
week, and the solid-tumor and non-Fc molecules do not.

| T-cell engager | Target | Format and route | Human half-life (days) |
|:---|:---|:---|---:|
| Elranatamab | BCMA | IgG, subcutaneous | 22 |
| Talquetamab | GPRC5D | IgG, subcutaneous | 8–12 |
| Mosunetuzumab | CD20 | IgG, intravenous | 6–11 |
| Glofitamab | CD20 | IgG 2:1, intravenous | 6–11 |
| Tarlatamab | DLL3 | half-life extended, intravenous | 5.7 |
| Teclistamab | BCMA | IgG, subcutaneous | 3.8 |
| HPN217 | BCMA | albumin-binding, intravenous | 2.8 |
| CB307 | PSMA | albumin-binding, intravenous | 3 |
| PF-06671008 | CDH3 | bispecific, intravenous | 1.2 |
| Tebentafusp | gp100 | ImmTAC, intravenous | 0.3 |
| Blinatumomab | CD19 | BiTE, continuous infusion | 0.09 |

: Published human half-lives. Every row except blinatumomab is from Table 1
of Pearce et al.; blinatumomab is the 2.11 hours in its label. ⚠️ Transcribed,
unverified. {tbl-colwidths="[24,12,34,30]"}

Two things separate the top of the table from the bottom, and neither is
target engagement. The long half-lives belong to molecules with an intact
Fc domain and therefore neonatal Fc receptor recycling, and the short ones
belong to formats without it. Blinatumomab at 54 kDa is below the renal
filtration threshold and is cleared in hours regardless of what it binds.
So the question is narrower than it first appears: **why does an
Fc-containing TCE, dosed at a fraction of its target pool, show the
half-life of an ordinary IgG?**

## 2. Four Explanations, and How Each Would Be Falsified

| # | Explanation | Would be falsified by |
|:--|:---|:---|
| 1 | The dose sits below the Michaelis constant, where saturable elimination is linear | A published $k_m$ comparable to, or below, the priming-dose peak concentration |
| 2 | The target sink is destroyed within the first cycle, so the nonlinearity has no time to show | A published model with a time-invariant nonlinear clearance and no decay term |
| 3 | Slow subcutaneous absorption rate-limits the terminal slope | An absorption rate constant fast relative to elimination, or the same flat profile intravenously |
| 4 | Sampling and assay quantitation limits hide the drop | Dense early sampling at a priming dose showing a log-linear decline |

: The four explanations and their falsifiers {tbl-colwidths="[5,55,40]"}

Explanations 1 and 2 both survive Sections 3 and 4. Explanation 3 survives
in one narrow case and is examined in Section 5. Explanation 4 is
untestable with what is public and is set out in Section 6 as a data
request rather than a finding.

## 3. Explanation 1: The Priming Dose Sits Below the Michaelis Constant

Odronextamab is the test case, because Kovalenko et al. published a model
with an explicit Michaelis-Menten elimination and reported $k_m$.

The structural model is a two-compartment disposition with parallel
first-order and Michaelis-Menten elimination, and the maximum velocity
declines with time since first dose:

$$
\frac{dV_{\max}}{dt} = -K_v\left(V_{\max} - V_{\max,\rm asym}\right),
\qquad
CL_{MM} = \frac{V_{\max}}{k_m + C}\,V_c
$$

with $V_{\max,0}$ = 2.93 mg/L/day, $k_m$ = 1.95 mg/L, $K_v$ = 1.09 /day,
$V_{\max,\rm asym}/V_{\max,0}$ = 0.262, linear $CL$ = 0.189 L/day,
$V_c$ = 4.99 L, $V_p$ = 4.42 L and $Q$ = 1.21 L/day.

```{r setup}
suppressMessages({
  library(rxode2); library(dplyr); library(tidyr)
  library(ggplot2); library(patchwork); library(xgxr)
})
xgx_theme_set()
knitr::opts_chunk$set(fig.width = 7, fig.height = 5)

odro <- rxode2({
  vmax  <- vmax_asym + (vmax0 - vmax_asym) * exp(-kv * time)
  cc    <- central / vc
  kmm   <- vmax / (km + cc)
  cltot <- cl + kmm * vc
  d/dt(central)    <- -kmm*central - cl/vc*central - q/vc*central + q/vp*peripheral
  d/dt(peripheral) <-  q/vc*central - q/vp*peripheral
})

par_odro <- c(cl = 0.189, q = 1.21, vc = 4.99, vp = 4.42,
              vmax0 = 2.93, km = 1.95, kv = 1.09)
par_odro <- c(par_odro, vmax_asym = unname(0.262 * par_odro[["vmax0"]]))
```

The approved regimen steps up 0.7 mg, 4 mg, 20 mg weekly and then holds at
160 mg. Peak concentrations after each of those doses, and the clearance
the model assigns to them, put every step-up dose on the flat part of the
Michaelis-Menten curve.

```{r regime}
#| fig-height: 4
#| fig-width: 6
#| fig-cap: "Total odronextamab clearance against concentration at the start of treatment, with the peak concentration of each step-up dose marked. The curve is flat below $k_m$, so the first three doses are eliminated by a saturable mechanism at a rate indistinguishable from first-order."
conc  <- 10^seq(-3.5, 2.5, length.out = 400)
curve <- tibble(conc, cltot = 0.189 + 2.93/(1.95 + conc)*4.99)
marks <- tibble(dose = c(0.7, 4, 20, 160, 320)) |>
  mutate(conc  = dose/4.99,
         cltot = 0.189 + 2.93/(1.95 + conc)*4.99,
         lab   = paste0(dose, " mg"))

ggplot(curve, aes(conc, cltot)) +
  geom_line(linewidth = 0.7) +
  geom_vline(xintercept = 1.95, linetype = 2, colour = "grey45") +
  geom_point(data = marks, size = 2.6, colour = "#C0392B") +
  ggrepel::geom_text_repel(data = marks, aes(label = lab), size = 3.2,
                           colour = "#C0392B", seed = 1) +
  annotate("text", x = 1.95, y = 0.25, label = "  km = 1.95 mg/L",
           hjust = 0, size = 3.2, colour = "grey35") +
  scale_x_log10() + scale_y_log10() + annotation_logticks() +
  labs(x = "Odronextamab concentration (mg/L)", y = "Total clearance (L/day)")
```

The 0.7 mg dose peaks at 0.14 mg/L, which is $k_m/14$. Simulating that
dose on its own, with target depletion switched off so that the only
nonlinearity left is the concentration dependence:

```{r linear-regime}
ev_single <- et(amt = 0.7, time = 0) |> et(seq(0, 14, by = 0.02))
no_deplete <- replace(par_odro, "kv", 0)
as_tibble(rxSolve(odro, no_deplete, ev_single)) |>
  filter(time %in% c(0.02, 0.5, 1, 2, 3, 7, 14)) |>
  transmute(`Day` = time,
            `Concentration (mg/L)` = signif(cc, 3),
            `Total CL (L/day)` = signif(cltot, 3)) |>
  knitr::kable()
```

Total clearance moves by 7% while concentration falls 900-fold. A
Michaelis-Menten elimination, simulated with its own published
parameters, produces a log-linear profile. This is the fourth phase of
TMDD in the Peletier and Gabrielsson classification, and Pearce et al.
name it as the reason TCEs retain linear pharmacokinetics at low doses
despite high-affinity binding.

**The verdict on explanation 1: it holds.** The priming dose is too small
to bend the curve. The sharp drop Andy expected requires a dose that
traverses the saturation boundary *during the observed profile*, and the
step-up schedule is built so that no single dose does that.

## 4. Explanation 2: The Target Sink Is Destroyed During Cycle 1

The nonlinearity does appear at the full dose, where concentration passes
$k_m$. By then the target is gone.

Simulating the full approved regimen, against two counterfactuals, shows
where the clearance goes.

```{r cl-time}
#| fig-height: 6
#| fig-cap: "Odronextamab under the 0.7/4/20/160 mg weekly regimen, as published and against two counterfactuals. Upper panel, concentration. Lower panel, total clearance. The published model loses 96% of its clearance over the first three weeks, from 7.2 to 0.30 L/day. With the target sink held fixed it loses only the part that saturation accounts for, and still clears 2.9 times faster than the published model at day 84."
ev <- et(amt = 0.7, time = 0) |> et(amt = 4, time = 7) |> et(amt = 20, time = 14) |>
      et(amt = 160, time = seq(21, 77, by = 7)) |> et(seq(0, 84, by = 0.02))

run <- function(pp, lab) as_tibble(rxSolve(odro, pp, ev)) |> mutate(case = lab)
sims <- bind_rows(
  run(par_odro, "as published"),
  run(replace(par_odro, "kv", 0), "target sink never depleted"),
  run(replace(par_odro, c("vmax0","vmax_asym"), c(0, 0)),
      "no target-mediated elimination"))

top <- ggplot(sims, aes(time, cc, colour = case)) +
  geom_line(linewidth = 0.6) + scale_y_log10() + annotation_logticks(sides = "l") +
  labs(x = NULL, y = "Concentration (mg/L)", colour = NULL)
bot <- ggplot(sims, aes(time, cltot, colour = case)) +
  geom_line(linewidth = 0.6) + scale_y_log10() + annotation_logticks(sides = "l") +
  labs(x = "Days since first dose", y = "Total clearance (L/day)", colour = NULL)

(top / bot) + plot_layout(guides = "collect") & theme(legend.position = "bottom")
```

The upper panel is what a reviewer of this study would have in front of
them. Three curves, each a near-straight line on a log axis within every
dosing interval, differing only in slope, with nothing that announces a
saturable mechanism. The lower panel holds the 25-fold change in clearance,
and clearance is not an observation.

Every published TCE population model of an approved drug encodes the same
decay, and each names target depletion or disease response as the cause.

| Drug | Clearance structure | At treatment start | Asymptote | Decay |
|:---|:---|---:|---:|:---|
| Odronextamab | Linear + Michaelis-Menten with time-declining $V_{\max}$ | ~5.6 L/day | 0.22 L/day | $K_v$ = 1.09 /day |
| Mosunetuzumab | $CL_{\rm base} \to CL_{ss}$, exponential | 1.08 L/day | 0.584 L/day | half-life 16.3 days |
| Teclistamab | $CL_1 + CL_2 e^{-K_{DES}t}$ | 0.996 L/day | 0.449 L/day | $K_{DES}$ = 0.0292 /day |
| Talquetamab | Time-independent + time-dependent linear | half-life 7.56 days | half-life 12.2 days | not transcribed |
| Epcoritamab | Quasi-steady-state TMDD, two-compartment | apparent CL/F 0.53 L/day at end of cycle 3 | not transcribed | concentration-driven |
| Glofitamab | Linear, two-compartment | 0.6–1 L/day | same | none |

: How each published model handles the target sink. ⚠️ Transcribed, unverified.
{tbl-colwidths="[16,32,18,15,19]"}

Glofitamab is the row to read against the other five. It is the one TCE
given after a mandatory 1000 mg obinutuzumab pretreatment, which depletes
CD20-bearing B cells a week *before* the first glofitamab dose, and its
published model is linear with no time-varying term. Remove the target
before the drug arrives and the target-mediated clearance is not there to
model.

Teclistamab's authors tested a quasi-steady-state TMDD parameterization
and a time-varying soluble BCMA covariate, and rejected both in favor of
the empirical exponential decay. Reading why those two failed is what
tells you whether the empirical form is a convenience or a necessity, and
it is entry 4 of the reading queue.

**The verdict on explanation 2: it holds, and it is why TMDD is
mis-filed.** A modeler who fits time-varying clearance has fit TMDD. The
word does not appear in the parameter table, so a reader scanning
parameter tables across this literature concludes that TCEs do not show
TMDD.

## 5. Explanation 3: Flip-Flop Kinetics After Subcutaneous Dosing

Andy's hypothesis is that slow subcutaneous absorption rate-limits the
terminal slope, so the apparent half-life reports absorption rather than
elimination. Two subcutaneous drugs have transcribed absorption rate
constants, and they answer differently.

```{r flipflop}
eff_thalf <- function(cl, vc, vp) log(2) * (vc + vp) / cl

tibble(
  Drug = c("Teclistamab", "Teclistamab", "Elranatamab"),
  When = c("treatment start", "steady state", "all times (linear model)"),
  `Absorption t1/2 (d)` = signif(log(2) / c(0.133, 0.133, 0.148), 3),
  `Elimination t1/2 (d)` = signif(c(eff_thalf(0.449 + 0.547, 4.13, 1.34),
                                    eff_thalf(0.449,         4.13, 1.34),
                                    eff_thalf(0.335,         4.25, 7.80)), 3)) |>
  mutate(`Rate-limiting step` =
           ifelse(`Absorption t1/2 (d)` > `Elimination t1/2 (d)`,
                  "absorption (flip-flop)", "elimination")) |>
  knitr::kable()
```

Teclistamab crosses the boundary during treatment. At treatment start
absorption is slower than elimination, so the terminal slope after an early
subcutaneous dose reports $K_a$; once the time-dependent clearance component
of Section 4 has decayed, elimination is the slower of the two and the
ordering has reversed. That crossing is a second reason teclistamab's
half-life lengthens with repeated dosing, working in the same direction as
the falling clearance.

Elranatamab does not cross it. Its absorption half-life of 4.7 days sits
against an elimination half-life near 25 days, so absorption is never
rate-limiting, and its published half-life of 22 days is elimination being
reported as elimination.

Three limits on how far the hypothesis carries.

1. **It cannot explain the intravenous drugs.** Mosunetuzumab, glofitamab
   and odronextamab have no absorption phase and show the same absence of
   visible nonlinearity. Explanation 3 is an additional flattening
   mechanism for some subcutaneous molecules, not the reason the class
   looks the way it does.
2. **It is contingent on where in treatment the sample was drawn.** The
   same drug is flip-flop in week 1 and not in week 8, because the quantity
   it is being compared against is itself changing. Any statement that a TCE
   does or does not show flip-flop kinetics has to name the time.
3. **Two cases is not a survey.** Epcoritamab reports a median time to
   maximum concentration of 4 days after the first full dose, consistent
   with slow absorption, but its absorption rate constant has not been
   transcribed here. Talquetamab uses a sequential zero- then first-order
   absorption, which does not reduce to a single $K_a$. Filling the
   absorption column of [PK parameters](pk-parameters.qmd) is what turns
   this section from two worked examples into a finding.

**The verdict on explanation 3: it holds for teclistamab early in
treatment, fails for elranatamab, and cannot apply to the intravenous
drugs.** It is a contributing mechanism with a narrow domain, not the
explanation.

## 6. Explanation 4: Sampling and the Quantitation Limit

Untestable from published summaries, and stated here so that it is not
mistaken for a settled negative.

Three features of how TCE pharmacokinetic data is collected would hide a
concentration drop that was there.

- **CD3 internalizes with a half-life near an hour.** The fastest
  target-mediated phase is over within a day. Cycle 1 sampling in these
  studies is typically predose, end of infusion, and then a small number
  of points across the following week.
- **Priming-dose concentrations sit near the assay limit.** The
  odronextamab simulation above puts the 0.7 mg profile between 0.14 and
  0.005 mg/L. Kovalenko et al. excluded 3.4% of post-dose records as below
  the limit of quantitation, and those records are disproportionately the
  early low-dose ones, which is exactly where the curvature would be.
- **The step-up schedule redoses before the profile resolves.** Each
  weekly dose interrupts the decline of the previous one.

What would settle it is a densely sampled single priming dose with a
low-limit assay, and the studies that could produce it are the ones
already run. A first-in-human dose escalation collects that data at the
lowest cohorts and reports a summary.

## 7. The Arithmetic of the CD3 Sink

Section 3 argues the priming dose is small relative to $k_m$. This section
asks the blunter version: how much drug does the CD3 pool actually consume?

```{r cd3}
avogadro  <- 6.022e23
cd3_per_T <- 5e4     # receptors per T cell
tcell_uL  <- 1500    # T cells per uL of blood
blood_L   <- 5
mw        <- 1.5e5   # g/mol
kint      <- 0.014   # per minute

tcells <- tcell_uL * 1e6 * blood_L
r_mol  <- tcells * cd3_per_T / avogadro
r_mg   <- r_mol * mw * 1000
vmax_mg_day <- r_mol * kint * 60 * 24 * mw * 1000

tibble(
  Quantity = c("T cells in blood",
               "CD3 receptors in blood (nmol)",
               "Drug for 1:1 occupancy of that pool (mg)",
               "CD3 internalization half-life (min)",
               "Maximum CD3-mediated elimination (mg/day)"),
  Value = c(format(tcells, digits = 3, scientific = TRUE),
            signif(r_mol * 1e9, 3),
            signif(r_mg, 3),
            signif(log(2)/kint, 3),
            signif(vmax_mg_day, 3))) |>
  knitr::kable()
```

Note the discrepancy in the internalization half-life. Pearce et al. give
the rate as 1.4% per minute and describe the corresponding half-life as
"about 36 min"; $\ln 2 / 0.014$ is 49.5 minutes. The 1.4% figure is used
above because it is the one their model consumes. Which of the two is the
measured quantity in the underlying source is an open check.

Against the epcoritamab regimen, which primes at 0.16 mg and reaches a full
dose of 48 mg weekly:

```{r cd3-doses}
tibble(Dose = c("0.16 mg priming", "0.8 mg intermediate", "48 mg full")) |>
  mutate(mg = c(0.16, 0.8, 48),
         `Multiples of the standing CD3 pool` = signif(mg / r_mg, 2),
         `Percent of one week's internalization capacity` =
           signif(100 * mg / (7 * vmax_mg_day), 2)) |>
  select(-mg) |>
  knitr::kable()
```

The priming dose is a few multiples of the standing receptor pool and a
rounding error against the pool's weekly turnover. The sink swallows it. The
full dose is several times the weekly capacity of the blood CD3 pool, which
is where the nonlinearity finally shows.

The calculation counts only circulating T cells. Whole-body T cells are
roughly fifty times the blood pool, and if a TCE reached all of them the
capacity would be near 100 mg/day and no clinical dose would ever saturate
CD3. Observed half-lives of a week say it does not. **How much of the tissue
T-cell pool is accessible to a TCE is the open question this calculation
exposes**, and it is the one to put to the Pearce model, which assumes blood
counts throughout.

## 8. Order of Work

1. **Check the transcribed parameters against the printed tables.** Every
   number in Sections 3, 4 and 5 is ⚠️ in [References](references.qmd).
   The odronextamab simulation reproduces a baseline total clearance of
   7.1 L/day where the paper reports a population median of 5.4 L/day, and
   whether that gap is the covariate model, between-subject variability,
   or a transcription error is unresolved.
2. **Fill the parameter table.** [PK parameters](pk-parameters.qmd) has
   eight drugs and four complete entries. Epcoritamab's quasi-steady-state
   parameters and talquetamab's time-dependent clearance are the two
   highest-value gaps, because they are the two published models with a
   structure nothing else here duplicates.
3. **Reproduce Figure 2 of Pearce et al.** Their simulation of half-life
   against CD3 affinity, from 1000 nM to 1 nM at a fixed 0.01 mg/kg dose,
   is the generic version of Section 3 and is reproducible from the
   equations in their methods.
4. **Test explanation 4 against one dense profile.** Any first-in-human
   report with individual concentration-time data at the lowest cohort
   would do.
5. **Write the note to Bill Denney.** Section 4 of
   [PK parameters](pk-parameters.qmd) lists what `nlmixr2lib` is missing
   and is already in a form that can be sent.
