---
title: "One structural model for the class"
subtitle: "A PK-only target-mediated model, calibrated across T-cell engagers, and what it cannot yet decide"
description: "A single structure with no CD3 arm, target in central and tissue, quasi-steady-state binding and drug-driven target killing. Calibrated to the published clearance changes for three T-cell engagers, with the identifiability problem that follows and the digitization plan that would resolve it."
categories: [pharmacometrics, pharmacokinetics, simulation, working document]
date: last-modified
execute:
  freeze: auto
  warning: false
  message: false
toc: true
toc-depth: 3
toc-location: left
html-math-method: mathjax
---

::: {.callout-important appearance="simple"}
**Nothing in this document has been fit to an observed concentration.** Every
result is a forward simulation from parameters transcribed out of published
tables, and the calibration in Section 3 targets a published *summary
statistic* rather than a profile. Section 5 states how much confidence that
supports, which is not much. The
[working specification](specification.qmd) carries the argument this model was
built to test, and [References](references.qmd) the status of every source.
:::

**Scope.** Molecules with an intact Fc domain and therefore neonatal Fc
receptor recycling. Blinatumomab, tebentafusp and the albumin-binding formats
clear on molecular size and are out.

## 1. The Structure

One model for the class, with no CD3 arm, no trimer, no cytokines and no
tumor growth.

$$
\begin{aligned}
\frac{dA_{\rm depot}}{dt} &= -k_a A_{\rm depot}\\
\frac{dA_c}{dt} &= k_a A_{\rm depot}-\frac{CL}{V_c}A_c
-\frac{Q_t}{V_c}A_c+\frac{Q_t}{V_t}A_t
-\frac{Q_p}{V_c}A_c+\frac{Q_p}{V_p}A_p
- k_{int}R_c V_c\frac{C_c}{K_{ss}+C_c}\\
\frac{dA_t}{dt} &= \frac{Q_t}{V_c}A_c-\frac{Q_t}{V_t}A_t
- k_{int}R_t V_t\frac{C_t}{K_{ss}+C_t}\\
\frac{dA_p}{dt} &= \frac{Q_p}{V_c}A_c-\frac{Q_p}{V_p}A_p\\
\frac{dR_j}{dt} &= k_{deg}\left(R_j^0-R_j\right)
-\left(k_{int}+k_{\rm kill}\right)R_j\frac{C_j}{K_{ss}+C_j},
\qquad j\in\{c,t\}
\end{aligned}
$$

Four drug compartments and two target pools. Target sits in the central
compartment and in a tissue compartment the drug must distribute into;
a second peripheral compartment carries no target and exists only to give
the disposition the shape antibodies actually have.

**Why CD3 is absent.** Section 9.1 of the specification ran the same model
with and without a CD3 arm at matched dose and target. The curvature ratios
were 7.5 and 7.0 at a priming dose and 11.3 and 11.3 at a saturating one.
The CD3 arm changes the profile by less than the width of the parameter
uncertainty, so carrying it costs four parameters and buys nothing.

**Why the target is killed and not only internalized.** $k_{int}$ removes
receptor with the drug bound to it; $k_{\rm kill}$ removes the cell carrying
it. Only the second can empty a pool faster than its own turnover, and
emptying the pool is what these drugs do.

```{r model}
suppressMessages({library(rxode2); library(dplyr); library(tidyr)
                  library(ggplot2); library(xgxr)})
xgx_theme_set()
knitr::opts_chunk$set(fig.width = 7, fig.height = 4.5)

uni <- rxode2({
  cc <- central/vc
  ct <- tissue/vt
  oc <- cc/(kss + cc)
  ot <- ct/(kss + ct)
  cltot <- cl + kint*rc*vc/(kss + cc) + kint*rt*vt/(kss + ct)
  rfrac <- (rc*vc + rt*vt)/(rc0*vc + rt0*vt)
  d/dt(depot)   <- -ka*depot
  d/dt(central) <-  ka*depot - cl/vc*central -
                    qt/vc*central + qt/vt*tissue -
                    qp/vc*central + qp/vp*periph - kint*rc*vc*oc
  d/dt(tissue)  <-  qt/vc*central - qt/vt*tissue - kint*rt*vt*ot
  d/dt(periph)  <-  qp/vc*central - qp/vp*periph
  d/dt(rc)      <-  kdeg*(rc0 - rc) - (kint + kkill)*rc*oc
  d/dt(rt)      <-  kdeg*(rt0 - rt) - (kint + kkill)*rt*ot
  rc(0) <- rc0
  rt(0) <- rt0
  f(depot) <- fbio
})

base <- c(cl = 0.20, vc = 5.0, vt = 4.0, vp = 3.0, qt = 1.2, qp = 0.6,
          kss = 1.95, kint = 1.0, kkill = 1.09, kdeg = log(2)/3,
          rc0 = 0, rt0 = 0, ka = 100, fbio = 1)

set_target <- function(p, total_mg, frac_tissue = 0.9) {
  p["rc0"] <- total_mg*(1 - frac_tissue)/p[["vc"]]
  p["rt0"] <- total_mg*frac_tissue/p[["vt"]]
  p
}
```

Every parameter above is shared across drugs except three: the route terms
$k_a$ and $F$, and the accessible target pool $R^0_cV_c+R^0_tV_t$. Section 3
asks whether moving that one pool is enough.

## 2. The Next Step: Why Teclistamab Reports 2.2-Fold and Odronextamab 26

Three candidates were open. Run them against each other in this structure,
under each drug's own approved regimen, and read the clearance at treatment
start against the clearance at day 120.

```{r regimens}
reg <- list(
  odronextamab = et(amt=0.7, cmt="central", time=0) |>
    et(amt=4, cmt="central", time=7) |> et(amt=20, cmt="central", time=14) |>
    et(amt=160, cmt="central", time=seq(21,119,by=7)),
  mosunetuzumab = et(amt=1, cmt="central", time=0) |>
    et(amt=2, cmt="central", time=7) |> et(amt=60, cmt="central", time=14) |>
    et(amt=60, cmt="central", time=21) |>
    et(amt=30, cmt="central", time=seq(42,119,by=21)),
  teclistamab = et(amt=4.8, cmt="depot", time=0) |>
    et(amt=24, cmt="depot", time=3) |> et(amt=120, cmt="depot", time=seq(7,119,by=7)))

cl_fold <- function(p, ev) {
  d <- as_tibble(rxSolve(uni, p, ev |> et(seq(0, 120, by = 0.02))))
  c(CL0 = d$cltot[d$time == 0.02], CLinf = d$cltot[d$time == 120])
}
sc_par <- function(p) { p["ka"] <- 0.133; p["fbio"] <- 0.718; p }
```

```{r three-candidates}
grid <- expand_grid(target_mg = c(1, 3, 14.6),
                    frac_tissue = c(0, 0.9),
                    route = c("IV", "SC"))
grid |>
  rowwise() |>
  mutate(r = list({ p <- set_target(base, target_mg, frac_tissue)
                    if (route == "SC") p <- sc_par(p)
                    cl_fold(p, reg$teclistamab) })) |>
  mutate(`CL0` = round(r[["CL0"]], 2), `CLinf` = round(r[["CLinf"]], 3),
         Fold = round(r[["CL0"]]/r[["CLinf"]], 1)) |>
  select(`Target (mg)` = target_mg, `Fraction in tissue` = frac_tissue,
         Route = route, CL0, CLinf, Fold) |>
  knitr::kable()
```

**Target burden is the lever and the other two are not.** Across a 15-fold
change in the accessible pool the reported fold change moves from 3.5 to 33.
Moving 90% of that pool into tissue changes it by less than 20%, and
switching the route from intravenous to subcutaneous changes it by about
30%, in the direction that makes the fold *larger* rather than smaller.

Note what this does not contradict. Section 9.5 of the specification found
target location decisive, and it was measuring a different thing: the
curvature of a single-dose profile, which is a shape. This is measuring the
ratio of two clearances under a full regimen, which is a number. Location
moves the shape and not the number.

## 3. Calibrating One Structure to Three Drugs

Hold everything fixed except the accessible target pool, and ask what pool
each drug's published fold change implies.

```{r calibration}
#| fig-height: 4
scan <- expand_grid(drug = names(reg), target_mg = 10^seq(log10(0.05), log10(30), length.out = 22)) |>
  rowwise() |>
  mutate(p = list({ p <- set_target(base, target_mg, 0.9)
                    if (drug == "teclistamab") p <- sc_par(p); p }),
         r = list(cl_fold(p, reg[[drug]])),
         fold = r[["CL0"]]/r[["CLinf"]]) |>
  ungroup()

published <- tibble(drug = names(reg), fold = c(25.7, 1.8, 2.2))

ggplot(scan, aes(target_mg, fold, colour = drug)) +
  geom_line(linewidth = 0.7) +
  geom_hline(data = published, aes(yintercept = fold, colour = drug),
             linetype = 2, linewidth = 0.5) +
  scale_x_log10() + scale_y_log10() + annotation_logticks() +
  labs(x = "Accessible target pool (mg of binding sites)",
       y = "Fold change in clearance, start to day 120", colour = NULL,
       caption = "Dashed lines are the published fold changes.")
```

| Drug | Published fold | Target pool that reproduces it |
|:---|---:|:---|
| Odronextamab | 25.7 | ~10 mg |
| Teclistamab | 2.2 | ~0.5 mg |
| Mosunetuzumab | 1.8 | ~0.3 mg **or** ~20 mg |

: One structure, one free parameter per drug {tbl-colwidths="[30,20,50]"}

A 20-fold difference in accessible target between lymphoma CD20 and myeloma
BCMA reproduces the published numbers. At 5×10⁴ sites per cell that is
8×10¹¹ cells against 4×10¹⁰, and at BCMA's lower density of 1.3×10⁴ sites
per cell the myeloma figure becomes 1.6×10¹¹ cells. Those are not absurd
numbers for nodal lymphoma against marrow-confined plasma cells in a heavily
pretreated population, which is the point at which a plausibility argument
stops being evidence.

## 4. The Identifiability Problem This Exposes

Mosunetuzumab has two answers, 60-fold apart. Its curve is not monotonic: the
fold change rises with target burden to about 9 and then collapses, because
the 30 mg every-three-weeks maintenance is a low enough dose intensity that a
large target pool is never fully destroyed, so the day-120 clearance stays
high and the ratio closes back up.

**A fold change in clearance does not identify a target pool.** It is one
number summarising two, and for a low-intensity regimen the map is
two-to-one. Every calibration in Section 3 inherits this, and the
odronextamab and teclistamab answers are single-valued only because their
weekly regimens happen to sit on the monotonic branch.

What separates the two mosunetuzumab solutions is the *shape* of the profile
during cycle 1, which the summary statistic throws away. That is the argument
for the next phase.

## 5. How Much Confidence Any of This Supports

Stated plainly, because the preceding four sections read more confidently
than the evidence warrants.

| Claim | Confidence | Why |
|:---|:---|:---|
| A saturable elimination is arithmetically first-order when $C\ll K_{ss}$ | **High** | Algebra, and it holds in any number of compartments |
| The published priming doses sit at $C/K_{ss}\le 0.07$ | **High** | Two transcribed numbers, one division; the transcription is unchecked |
| This structure is *capable* of spanning 1.8 to 26-fold on one parameter | **High** | Demonstrated in Section 3 |
| Target burden is what actually differs between these drugs | **Low** | A structure flexible enough to span the range by construction is not evidence that it spans it for the right reason |
| Target location decides the single-dose curvature | **Moderate** | It survives every perturbation tried, but $Q_t$ and the central/tissue split are assumed, not estimated |
| Any of this explains the observed data | **Low** | Nothing here has met an observed concentration |

: Confidence by claim {tbl-colwidths="[42,16,42]"}

Two results from this project so far bear on how much weight to put on
simulation alone, and they point the same way. The round-trip in Section 9.4
of the specification falsified the hypothesis it was built to confirm. The
standing-CD3-sink explanation in Section 4a was rejected by the first
simulation that tested it. Both were plausible before they were run, and
both were wrong. The reasonable prior on the remaining untested claims is the
same.

**The honest summary.** There is one arithmetic result that is certain and
one structure that can reproduce the published summary statistics. There is
no evidence yet that it reproduces them for the right reason, because it has
never been shown a concentration-time profile.

## 6. What Would Settle It

Fit, or overlay, against digitized profiles.

### 6.1 The Digitization Target

One table, one row per published figure, with the dose level and the
sampling window recorded so that the cycle-1 profiles can be told apart from
the steady-state ones.

| Source | Figure | What it would constrain |
|:---|:---|:---|
| Kovalenko 2026, odronextamab | Concentration-time by B-NHL subtype, cycles 1–3 and long-term follow-up | The one published profile set that spans a 10,000-fold dose range; separates the two mosunetuzumab-style branches directly |
| Bender 2024, mosunetuzumab | Observed versus predicted by dose group | Resolves the 0.3 mg against 20 mg ambiguity in Section 3 |
| Miao 2023, teclistamab | Concentration-time by route and dose, Figs 2 and 3 | The subcutaneous absorption and the myeloma target pool together |
| Pearce 2025 | Table 1 half-lives, 29 molecules | Already transcribed; the out-of-sample check on any fitted structure |
| CD20×CD3 review | Cynomolgus TMDD dose thresholds | The preclinical arm, where the elbow is reported visible |

: Digitization queue, ordered by what each resolves
{tbl-colwidths="[26,34,40]"}

### 6.2 The Sequence

1. **Digitize odronextamab first.** It is open access, its dose range is the
   widest published, and its model is the only one with a fitted $k_m$ to
   check the digitization against.
2. **Overlay before fitting.** Run this structure at the calibrated target
   pool and plot it over the digitized points. An overlay that is visibly
   wrong tells you more than a fit that converges.
3. **Fit only what the data can carry.** With digitized medians rather than
   individual data, the estimable set is small: the target pool, $K_{ss}$,
   and $k_{\rm kill}$, with disposition fixed to the published values.
4. **Then mosunetuzumab**, as the test of whether the ambiguity in Section 4
   closes.
5. **Then teclistamab**, which adds the subcutaneous absorption and is where
   the structure is most likely to break.

### 6.3 What Would Falsify the Structure

Written down now, before any data is fitted.

- A profile whose cycle-1 decline is *slower* than the structure predicts at
  the calibrated target pool, which would mean the accessible target is
  smaller still and something else carries the clearance.
- A drug needing a target pool outside the range its receptor density and
  disease burden allow.
- Any drug requiring a different $K_{ss}$ than its published binding affinity
  supports by more than about tenfold.
- Two drugs against the same antigen needing target pools that differ by more
  than their populations' tumor burdens differ.
