---
title: "Theophylline: data, model, and simulation"
subtitle: "Demo page — observed data against a fitted one-compartment model"
description: >
  The theophylline single-dose dataset plotted, a one-compartment model with
  first-order absorption fitted per subject, and the simulation from that fit
  compared against the observed data.
categories: [GenAI, demo]
date: last-modified
execute:
  freeze: true
  warning: false
  message: false
toc: true
---

::: {.callout-note appearance="simple"}
The model here is fitted with base R's `nls()`, not with `nlmixr2`. On the
machine this was built on, `rxode2` cannot compile a model — see the note at
the foot of the page. The intended `nlmixr2` version is a few lines and is
given there.
:::

```{r setup}
library(ggplot2)
library(dplyr)
theme_set(theme_bw(base_size = 12))
```

## The data

`theo_sd` is the theophylline single-dose study: 12 subjects, one oral dose,
concentrations followed for 24 hours.

```{r data}
theo <- nlmixr2data::theo_sd
obs  <- theo |> filter(EVID == 0)
dose <- theo |> filter(EVID != 0) |> select(ID, AMT) |> distinct()

obs |>
  ggplot(aes(TIME, DV, group = ID)) +
  geom_line(alpha = 0.4) +
  geom_point(size = 1) +
  labs(x = "Time (h)", y = "Concentration (mg/L)",
       title = "Observed theophylline concentrations, 12 subjects")
```

## The model

One compartment, first-order absorption, fitted to each subject:

$$C(t) = \frac{D\,k_a}{V\,(k_a - k_e)}\left(e^{-k_e t} - e^{-k_a t}\right)$$

```{r fit}
one_cmt <- function(t, D, ka, ke, V) {
  D * ka / (V * (ka - ke)) * (exp(-ke * t) - exp(-ka * t))
}

fits <- lapply(split(obs, obs$ID), function(d) {
  D <- dose$AMT[dose$ID == d$ID[1]]
  f <- nls(DV ~ one_cmt(TIME, D, ka, ke, V), data = d,
           start = list(ka = 1.5, ke = 0.08, V = 0.5),
           control = nls.control(maxiter = 200, warnOnly = TRUE))
  cf <- as.list(coef(f))
  data.frame(ID = d$ID[1], ka = cf$ka, ke = cf$ke, V = cf$V, D = D)
})
pars <- do.call(rbind, fits)
knitr::kable(pars, digits = 3, caption = "Fitted parameters by subject")
```

## Simulation against the observed data

```{r simulate}
grid <- do.call(rbind, lapply(seq_len(nrow(pars)), function(i) {
  p <- pars[i, ]
  t <- seq(0, max(obs$TIME), length.out = 200)
  data.frame(ID = p$ID, TIME = t,
             SIM = one_cmt(t, p$D, p$ka, p$ke, p$V))
}))

ggplot() +
  geom_line(data = grid, aes(TIME, SIM, colour = "Simulated")) +
  geom_point(data = obs, aes(TIME, DV, colour = "Observed"), size = 1) +
  facet_wrap(~ ID) +
  scale_colour_manual(values = c(Observed = "black", Simulated = "#2c7fb8")) +
  labs(x = "Time (h)", y = "Concentration (mg/L)", colour = NULL,
       title = "Observed data and the fitted model, by subject") +
  theme(legend.position = "bottom")
```

```{r dv-pred}
obs_pred <- obs |>
  left_join(pars, by = "ID") |>
  mutate(PRED = one_cmt(TIME, D, ka, ke, V))

ggplot(obs_pred, aes(PRED, DV)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey60") +
  geom_point(alpha = 0.7, size = 1) +
  labs(x = "Predicted (mg/L)", y = "Observed (mg/L)",
       title = "Observed against predicted")
```

## What blocked the nlmixr2 version

`rxode2` 5.0.2 and `nlmixr2` 5.0.0 are installed and the model below is
accepted, but building it fails at the compile step with *"error building
model — please make sure you have a working C compiler set up"*. `clang`,
`make` and the Command Line Tools are present, and `R CMD SHLIB` compiles a
plain C file, so R's toolchain works; the failure is inside `rxode2`, which
does not reach the point of writing its generated `.c` file.

```r
one.cmt <- function() {
  ini({
    tka <- 0.45; tcl <- 1; tv <- 3.45
    eta.ka ~ 0.6; eta.cl ~ 0.3; eta.v ~ 0.1
    add.sd <- 0.7
  })
  model({
    ka <- exp(tka + eta.ka)
    cl <- exp(tcl + eta.cl)
    v  <- exp(tv  + eta.v)
    linCmt() ~ add(add.sd)
  })
}
fit <- nlmixr2(one.cmt, theo_sd, est = "focei")
```
